Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/discussion.tex
Minor paper edits.  A few TODOs inserted.

%This analysis suggests 
Our approach has the ability to uncover systematic differences in dynamic behaviors or roles among subsets of nodes.
% across a variety of data sets.
In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
In particular, we show that our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are both special cases of our model.
Though prediction is not the main focus of the model, the results provide some evidence that having latent structure (i.e. K>1) can lead to improvements in predictive power. Our prediction results show that it is easy to overfit with this model. If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model. % (TODO: confirm this).  
However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit. This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model. A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.

>>>>

%This analysis suggests 
Our approach has the ability to uncover systematic differences in dynamic behaviors among subsets of nodes.
% across a variety of data sets.
In Section \ref{sec:experiments} we show the model has improved predictive accuracy over baseline methods on real data with respect to ranking tasks and the likelihood of unobserved data.
In particular, we show that our proposed approach leads to improved predictive accuracy when compared to both (a) relational event models that lack latent clusters, and (b) stochastic blockmodels for count data, both of which are both special cases of our model.
Though prediction is not the main focus of the model, the results provide some evidence that having latent structure (i.e. K>1) can lead to improvements in predictive power. Our prediction results show that it is easy to overfit with this model. If we restrict to K=2 on these small data sets, we systematically outperform the K=1 model. {\red(TODO: confirm this)}
  However, as K increases beyond 2 (e.g., for the unrestrained CRP prior), the model is prone to overfit. This is not surprising, given that the number of fitted parameters scales as the square of the number of components K in the model. A natural future direction worth exploring for this model is a class of priors that are more resistant to overfitting.

### 2012-06-01 docs/experiments.tex
Add TODOs in results section.
Add another placeholder and update TODO file.
Minor paper edits.  A few TODOs inserted.

Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.

@@

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{p,1,3}$ and $\beta_{p,3,1}$.
Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.

@@

We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood of held-out data and recall on held out data.
Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.  I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.

@@

%\input{../figs/results-recall.tex}

>>>>

A variety of real world datasets are used to explore the efficacy of the model.  Each of the following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.

@@

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
The two panels on the left show the variation in $s_0(t_M,i,j)$, $s_1(t_M,i,j)$ and $s_3(t_M,i,j)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
Similarly, block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data and our inferences from the parameter estimates are sensible given the observed statistics.

@@

We evaluate the predictive ability of the fitted models by comparing models based on the loglikelihood and recall for held-out data.
Each data set is first split into a training set and a test set, and the loglikelihood of the test set is computed sequentially using Equation \ref{eqn:llk} where $\beta$ is set to be the mean from posterior samples given the training data.\footnote{This approach makes sense when the latent class assignments are not changing.  I will change this so that we averaging across predictions made from single draws from the posterior $\beta^{(i)}, z^{(i)}$.}
We compute both the relational event model likelihood  (\texttt{rem}) as given in Equation \ref{eqn:llk}  and the multinomial likelihood (\texttt{mult}) given in Equation \ref{eqn:multllk}.
The latter only measures a model's predictive performance for \emph{what} occurs next, while the former also measures the ability to predict \emph{when} it occurs.

@@

[TODO: Recall table.]

[TODO: Mention that we cap the value of $K$]

[TODO: Discuss results.]
%\input{../figs/results-recall.tex}

### 2012-06-01 docs/simulation.tex
Minor paper edits.  A few TODOs inserted.

Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
Intensities for reciprocal actions among nodes in block 1 are $e^3$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).

>>>>

Though the dyad counts for the synthetic data set suggest a stochastic blockmodel (as seen in Figure  \ref{fig:syncounts}),  the center plot shows each block has empirical differences in their dynamics.
Intensities for reciprocal actions among nodes in block 1 are $e^3 \approx 20$ times greater, intensities for turn-taking actions among nodes in group 2 are $e^2$ times greater, and intensities for dyadic interactions between the two groups have a multiplicative effect of $e^{-1}$ and thus occur less often.
Fitting the model with $K=2$ has similar  predictive accuracy as the true model (see Table \ref{tab:results}), recovers the true latent classes, and the posterior credible intervals of the parameters cover the true parameter values (see Figure \ref{fig:syncounts}c).
